Measuring neutron star tidal deformability with Advanced LIGO: a Bayesian analysis of neutron star - black hole binary observations

### 2016-09-21 Paper/paper.tex
mp comments v2

\usepackage[colorlinks, pdfborder={0 0 0}, plainpages=false]{hyperref}
\usepackage{graphicx}
\usepackage{xspace}
\usepackage[usenames,dvipsnames]{color}
\usepackage{amssymb}
\usepackage[dvipsnames]{xcolor}
\usepackage{placeins}
% \usepackage[switch,columnwise]{lineno} 
\usepackage{lineno}
\interfootnotelinepenalty=10000

@@

% \red{
% Text about CBCs. How was each type observed so far?
% }
Coalescing compact binaries of stellar-mass black holes (BH) and/or
neutron stars (NS) are the primary targets for the second
generation GW detectors~\cite{Timmes:1995kp,Fryer:1999mi,RevModPhys.74.1015,
2010ApJ...714.1217B,2010ApJ...715L.138B,Dominik:2014yma,Belczynski:2006zi,
2012ApJ...749...91F,
Wex:1998wt,1991ApJ...379L..17N,Mandel:2015spa,Abbott:2016nhf}.
A binary system of black holes was recently observed by 
aLIGO~\cite{LIGOVirgo2016a}. Previously, stellar-mass black holes had only 
been observed by inference in mixed binaries with stellar companion (through
electromagnetic observations of the companion)~\cite{Lewin2010,
Remillard:2006fc,Fragos:2010tm}.
Neutron stars, on the other hand, have had numerous sightings. Thousands of
electromagnetically emitting neutron stars, or pulsars, have been 
documented~\cite{Manchester:2004bp},
in varied situations: as radio pulsars~\cite{Lattimer:2012nd,Manchester:2004bp},
in binary systems with a stellar companion~\cite{1971ApJ...169L..23M,
Bond:2002eh,Lattimer:2012nd,Manchester:2004bp},
and in neutron star binaries~\cite{Hulse:1975uf,Taylor:1982wi,
Weisberg:2010zz,Lattimer:2012nd,Manchester:2004bp}.
Mixed binaries of black holes and neutron stars, is an astrophysically
interesting class of systems~\cite{Wex:1998wt,
1991ApJ...379L..17N,Janka1999,Fryer:2015jpa}, that has not yet been detected.
We expect to observe $\mathcal{O}(10)$ mixed binaries per year with
aLIGO~\cite{Abadie:2010cf}.

@@

Another question that NSBH mergers can help answer is `what is the 
nature of matter at nuclear densities', i.e. `what is the EoS of NS material'?
This is what we will focus on in this paper.
% 
During the course of early inspiral, the tidal field of the BH produces a
deformation in its companion NS. The quadrupolar moment of the star associated
with this deformation also depends on its material properties, through an EoS-dependent
tidal deformability parameter $\lambdans$. This induced quadrupolar moment
changes over the orbital time-scale, resulting in the emission of GWs in {\it
coherence} with the orbital waves.
These waves draw more energy from the orbit and increase the inspiral rate (as
compared to an equivalent BBH)~\cite{Flanagan2008}.
% The coherence of GW emission between stellar and orbital waves drains energy 
% more rapidly from the orbit and increases the inspiral rate (compared to a
% BBH)~\cite{Flanagan2008}.
% 
% We know the leading and next-to-leading order terms in post-Newtonian (PN)
% theory that capture this effect~\cite{Vines2011}, entering binary phasing at
% $5$PN order.
% 
Closer to merger, the strong tidal field of the BH can disrupt the NS. The
quadrupolar moment of the disrupted binary system falls monotonically over a
millisecond time-scale~\cite{Kyutoku:2010zd,Lackey:2013axa,Lovelace:2013vma,
Foucart:2015a,Pannarale:2015jia}, resulting in the damping of GW amplitude.
% 
This penultimate stage also depends strongly on the internal structure and energy
transport mechanism of the NS, and carries the strongest tidal signature in the
GW spectrum~\cite{Foucart:2014nda,Deaton2013}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Techniques}\label{s1:techniques}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% #################
% THESE FIGURES ARE FOR THE NEXT SECTION
\begin{figure*}
\centering 
\textbf{Illustrative posterior probability distributions for mass-ratio $\eta$,}\par
\textbf{at different SNR values (shown in legend).}\par\medskip
\includegraphics[width=1.8\columnwidth]{plots/SingleSystemEta_q4_0_mc2_25_chi0_50}
\caption{We show here probability distributions for mass ratio $\eta$ as measured
for the same signal at different SNRs. The intrinsic parameters of the source
are: $q = \mbh/\mns = 5.4M_\odot/1.35M_\odot = 4$, $\chibh=+0.5$, and $\lambdans=2000$;
and the signal is injected at SNRs $\rho=\{20,30,50\}$ (left to right). The templates
{\it ignore} tidal effects.
% 
In each panel: the dashed red line marks the median or {\it measured} value
$\eta^\mathrm{Median}$, while the dashed green line show the true value
$\eta^\mathrm{Injected}$. The darker shading shows
the recovered $90\%$ credible interval for $\eta$, $(\Delta\eta)^{90\%}$.
% 
Comparing systematic and statistical errors, we find that:
at $\rho=20$, $\eta$ measurement is dominated by statistical
errors; at $\rho=30$, the two become comparable; and 
for louder signals ($\rho\simeq50$), the systematic errors dominate.
}
\label{fig:SingleSystemEtaPDFvsSNR}
\end{figure*}

% ###############################
% In this section we summarize various technical aspects of this paper. The model
% used to obtain tidal waveforms is described in Sec.~\ref{s2:waveforms}. The 
% process of inferring source parameters from a GW signal is summarized in 
% Sec.~\ref{s2:bayesian}. The process of generating and combining multiple events
% is instead detailed in the corresponding results section,
% Sec.~\ref{s2:astro_multiple}.

@@

\textcolor{blue}{%
Credible intervals $(\Delta X)^{90\%}$ showing the precision with which
$X=\{\mchirp,\eta,\chibh\}$ can be measured, are presented in
Appendix~\ref{as1:nontidalerrors}.
}
We remind ourselves that this {\it precision} is only meaningful so long as the
measurement is {\it accurate} to begin with. Therefore, we define $\arr_X$ as
the ratio between systematic and statistical errors associated with the
measurement of parameter $X$,
\begin{equation}\label{eq:arr}
\arr_X = \dfrac{(X^\mathrm{Median} - X^\mathrm{Injected})}{(\Delta X)^{90\%}},
\end{equation}
in order to compare the relative magnitude of both. Only when
$|\arr_X| < 1$ can we ignore tidal effects in our templates
without hampering the measurement of non-tidal parameters from NSBH signals.
% 
We start with calculating $\arr_{\mchirp}$ as a function of various source
parameters and show it in Fig.~\ref{fig:TN_chirpMassBias_vs_Lambda_SNR}.
$\mchirp$ is the leading order
mass combination that affects the GW strain emitted by compact binaries as they
spiral in, and is therefore determined the most precisely. We notice
immediately that for $\rho\leq 30$ the systematics are well under control
and so we can obtain reliable chirp mass estimates for NSBH signals using BBH
templates.
% 
For louder and less likely SNRs ($\rho\simeq 50$), we find that
$\arr_{\mchirp}$ can become comparable to unity, but only if the BH has
prograde spin $\chibh\gtrsim 0.4$, {\it and} the true NS tidal deformability
is large enough, s.t. $\lambdans \gtrsim 1000$.
% 
We can therefore conclude that only for very loud signals, with
$\rho\gtrsim 50-70$, will the inclusion of tidal terms in our template
models improve our $\mchirp$ estimation. For lower SNRs, inclusion of new
physical content in templates will instead get washed out by detector noise.
% 
In addition to everything, we also note that $\arr_{\mchirp}\geq 0$ always,
i.e. $\mchirp$ is always being over-estimated. This is to be expected since
the tidal deformation and disruption of NSs close to merger reduces GW signal
power at high frequencies, making the resulting signal resemble a BBH signal of
lower merger frequency - and therefore of higher mass.

@@

In the previous section, we showed that the effects of the tidal deformation of
NSs by their companion BHs become discernible in
the GW spectrum under certain favorable conditions, including (a) BH mass is
sufficiently small, (b) BH spin is positive aligned, i.e. $\chibh\gtrsim +0.4$,
(c) the NS is not very compact, with $\lambdans\gtrsim 1000$, and (d) the
source location and orientation are such that its GW SNR $\gtrsim 30$.
% 
Both condition (a) and (b) enhance the tidal distortion of the star and increase
the number of orbits the system goes through at small separation, where the
differences between NSBH and BBH signals are maximal.
% 
Conditions (a)-(c) also reduce the onset frequency of the disruption of the NS,
allowing for it to happen earlier in the orbit. 
% 
We expect that these conditions are also the ones which should maximize the
likelihood of {\it measuring} tidal effects in NSBH signals. In this section,
%
%In addition, a SNR of $30$ corresponds to a $1.35+5M_\odot$ binary with 
%$\chibh = 0.4$ at a luminosity distance of $XXX$~Mpc, if optimally aligned
%with LIGO ($XXX$~Mpc if the binary subtends a $45^o$ angle with LIGO's line of
%sight.
%
%These conditions are strong, but not unrealistic. For e.g., a detection rate
%of $20$ NSBHs a year implies $1-2$ of them will have SNRs$\gtrsim 30$, since
%the number of expected events scales as $1/\rho^2$ if we assume that sources
%are distributed uniformly in spacial volume.
%
we turn the question around to ask: under similarly favorable circumstances,
can we gain insights about the internal structure of neutron stars from GW
observations?

@@

In the previous section, we showed that single observations of NSBH coalescences
at moderate SNRs have little information about the internal structure of neutron
stars that will be accessible to Advanced LIGO at its design sensitivity. We
expect all neutron stars to share the same equation of state, and hence the same
$\lambdans(\mns)$. In addition, we know that the mass distribution of (most) NSs
that have not been spun up to millisecond periods (which are what we focus on in 
this paper, by approximating $\chins\approx 0$) is narrowly peaked around
$\sim 1.35M_\odot$~\cite{Kiziltan2013}. Therefore, information from multiple NSBH
observations can be combined to improve our estimation of $\lambdans$ for the
population. This is what we explore in this section, within a fully Bayesian 
framework.

@@

multiple {\it identical} sources at realistic SNRs. Let us consider the case

@@

In order to generate each population, the first step is to fix
the NS properties: (i) NS mass $\mns=1.35M_\odot$, (ii) NS spin $\chins=0$
and (iii) NS tidal deformability $\lambdans=$ fixed value chosen from
$\{500,800,1000,1500,2000\}$. Next, we generate events, by sampling BH mass
(uniformly) from $\mbh\in[3M_\odot,6.75M_\odot]$, BH spin (uniformly)
from $\chibh\in[0, 1]$, source location uniformly in spatial volume,
and its inclination from $\iota\in[0,\pi]$. Note that we only
sample positive aligned BH spins, since binaries with anti-aligned spins have
very little information to add at realistic SNRs, as shown in 
Fig.~\ref{fig:TT_LambdaCIWidths90_0_Lambda_SNR}. This further restricts the
sub-space of all NSBH observations that our populations span, and will need to
be take into account when the number of observations required for tidal
measurements is related to detector observation time.
% 
As noted previously, we also restrict events to have
$\rho\geq\rho_\mathrm{min}$, where $\rho_\mathrm{min}=10$.
Finally, we repeat the process till we have an ordered set of $N$ events.
% 
Since we want to analyze not just a single realization of an astrophysical
population, but an ensemble of them, we make an additional approximation to
mitigate computational cost. Complete Bayesian parameter estimation is
performed for a set of simulated signals whose parameters are the vertices
of a regular hypercubic grid (henceforth ``G'') in the space of
$\{q\}\times\{\chibh\}\times\{\rho\}$, with each sampled at $q=\{2,3,4,5\}$,
$\chibh=\{-0.5,0,0.5,0.75\}$, and $\rho=\{10,20,30,50,70\}$.
All events in each population draw are substituted by their
respective nearest neighbours on the grid G.

@@

First, we study the effects of not including tidal terms in our search
templates while characterizing NSBH signals. We expect that the waveform
template that best fits the signal would compensate for the reduced number of
degrees of freedom in the template model by moving away from the true
parameters of the binary. This should result in a {\it systematic} bias in 
the recovered values of non-tidal source parameters, such as its masses 
and spins. In order to quantify it, we inject tidal signals into zero noise,
and perform a Bayesian parameter estimation analysis on them using templates
{\it without} tidal terms.
% 
We use the LEA+ model~\cite{Lackey:2013axa} to produce tidal waveforms that
incorporate the effect
of NS distortion during inspiral, and of its disruption close to merger. Our
injected signals sample the region of NSBH parameter space where NS disruption
prior to binary merger is likely {\it and} can be modeled using LEA+. Their
parameters are given by combinations of $q=\mbh/\mns=\{2,3,4,5\},
\chibh=\{-0.5, 0, 0.5, 0.75\}$ and $\lambdans=\{500,800,1000,1500,2000\}$.
Other parameters, such as source location and orientation, that factor out of
$h(t)$ as amplitude scaling are co-sampled by varying $\rho=\{20,30,50,70\}$.

@@

% \prayush{\bf Sec 5 summary}\\
% \prayush{\bf Tidal measurements versus observation time - in how many years can
% we constrain $\lambdans$ reasonably?}\\
% % 
Third, knowing that single observations can furnish only so much information
about the NS equation of state, we move on to investigate how well do we do with
multiple signals. In order to quantify how $\lambdans$ measurement improves
with the number of observed events $N$, we generate populations of NSBH signals
and combine the information extracted from each event.
% 
The population generation procedure is as follows. The neutron star mass is
held fixed at $1.35M_\odot$, its spin at $\chins=0$, and its tidal
deformability is fixed to each of $\lambdans=\{500,1000,1500,2000\}$. Black hole
mass is sampled uniformly from the range $[2,5]\times 1.35=[2.7, 6.75]M_\odot$,
and spin from $\chibh\in[0,1]$. As before, our parameter choice here is given
by the intersection set of the mass range that allows for neutron star disruption
and the range supported by LEA+~\cite{Foucart2012,Foucart:2013a,Lackey:2013axa}.
In order to keep the computational cost reasonable, we make an additional
approximation. For every population generated, we replace each event's
parameters by their nearest neighbor on the uniform grid G, which has vertices
at: $q=\{2,3,4,5\}\times\chibh=\{-0.5,0,0.5,0.75\}\times\lambdans=\{500,800,
1000,1500,2000\}\times\rho=\{10,20,30,50,70\}$.
% 
This way, we only have to run full Bayesian parameter estimation analysis on
this fixed set of signals. 
% 
There are two sources of error that will enter the deductions we will make from
a a single population generated in the manner described above. First, since the
injection parameters are pushed to their nearest neighbor on a grid, we would 
find discrete jumps in $\lambdans$ errors as a function of $N$. And second, an
individual population is one particular realization of a stochastic process and
could have excursions that may never be found in another population. To
account for both of these limitations, we generate an ensemble of populations,
and conservatively combine information from all of them\footnote{See 
Sec.~\ref{s1:multiple_observations} for further details.}.

@@

Finally, we note that a more recent work that improves upon LEA+ has been
published~\cite{Pannarale:2015jka}, but it
provides only an amplitude model which needs to be (in future) augmented with a
compatible phase model. In addition, accuracy of the underlying numerical
simulations used to calibrate the waveform model used here have not been
verified against independent codes so far. It is therefore difficult to assess
the combined modeling error and its effect on our results. Our results here
are, therefore, limited by the limitations of our waveform model, and presented
with this caveat. However, we expect the combined effect of modeling errors to
{\it not} affect our qualitative conclusions.

>>>>

\usepackage[colorlinks, pdfborder={0 0 0}, plainpages=false]{hyperref}
\usepackage{graphicx}
\usepackage{xspace}
\usepackage[usenames,dvipsnames]{color}
\usepackage{amssymb}
\usepackage[dvipsnames]{xcolor}
\usepackage{placeins}
\usepackage[utf8]{inputenc}
% \usepackage[switch,columnwise]{lineno} 
\usepackage{lineno}
\interfootnotelinepenalty=10000

@@

% \red{
% Text about CBCs. How was each type observed so far?
% }
Coalescing compact binaries of stellar-mass black holes (BH) and/or
neutron stars (NS) are the primary targets for the second
generation GW detectors~\cite{Timmes:1995kp,Fryer:1999mi,RevModPhys.74.1015,
2010ApJ...714.1217B,2010ApJ...715L.138B,Dominik:2014yma,Belczynski:2006zi,
2012ApJ...749...91F,
Wex:1998wt,1991ApJ...379L..17N,Mandel:2015spa,Abbott:2016nhf}.
A binary system of black holes was recently observed by 
aLIGO~\cite{LIGOVirgo2016a}. Previously, stellar-mass black holes had only 
been observed by inference in mixed binaries with stellar companion (through
electromagnetic observations of the companion)~\cite{Lewin2010,
Remillard:2006fc,Fragos:2010tm}.
Neutron stars, on the other hand, have had numerous sightings. Thousands of
electromagnetically emitting neutron stars, or pulsars, have been 
documented~\cite{Manchester:2004bp},
in varied situations: as radio pulsars~\cite{Lattimer:2012nd,Manchester:2004bp},
in binary systems with a stellar companion~\cite{1971ApJ...169L..23M,
Bond:2002eh,Lattimer:2012nd,Manchester:2004bp},
and in binary neutrons (BNS)~\cite{Hulse:1975uf,Taylor:1982wi,
Weisberg:2010zz,Lattimer:2012nd,Manchester:2004bp}.
Mixed binaries of black holes and neutron stars, is an astrophysically
interesting class of systems~\cite{Wex:1998wt,
1991ApJ...379L..17N,Janka1999,Fryer:2015jpa}, that has not yet been detected.
We expect to observe $\mathcal{O}(10)$ mixed binaries per year with
aLIGO~\cite{Abadie:2010cf}.

@@

Another question that compact object mergers can help answer is `what is the 
nature of matter at nuclear densities', i.e. `what is the EoS of NS material'?
A large fraction of past work aimed at measuring NS matter effects from GW
signals has consisted of inquiries about BNSs~\cite{Lee1999a,Lee1999b,Lee2000,
oechslin:07,Read:2008iy,Markakis:2010mp,Markakis:2011vd,stergioulas:11,
East:2011xa,Lackey2014,Wade:2014vqa,Bauswein:2014qla}. In this paper, we will
instead focus on NSBHs.
% 
During the course of early inspiral, the tidal field of the BH produces a
deformation in its companion NS. The quadrupolar moment of the star associated
with this deformation also depends on its material properties, through an EoS-dependent
tidal deformability parameter $\lambdans$. This induced quadrupolar moment
changes over the orbital time-scale, resulting in the emission of GWs in {\it
coherence} with the orbital waves.
These waves draw more energy from the orbit and increase the inspiral rate (as
compared to an equivalent BBH)~\cite{Flanagan2008}.
% The coherence of GW emission between stellar and orbital waves drains energy 
% more rapidly from the orbit and increases the inspiral rate (compared to a
% BBH)~\cite{Flanagan2008}.
% 
% We know the leading and next-to-leading order terms in post-Newtonian (PN)
% theory that capture this effect~\cite{Vines2011}, entering binary phasing at
% $5$PN order.
% 
Closer to merger, the strong tidal field of the BH can disrupt the NS. The
quadrupolar moment of the disrupted binary system falls monotonically over a
millisecond time-scale~\cite{Kyutoku:2010zd,Lackey:2013axa,Lovelace:2013vma,
Foucart:2015a,Pannarale:2015jia}, resulting in the damping of GW amplitude.
% 
This penultimate stage also depends strongly on the internal structure and energy
transport mechanism of the NS, and carries the strongest tidal signature in the
GW spectrum~\cite{Foucart:2014nda,Deaton2013}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Techniques}\label{s1:techniques}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% #################
% THESE FIGURES ARE FOR THE NEXT SECTION
\begin{figure*}
\centering 
\textbf{Illustrative posterior probability distributions for mass-ratio $\eta$,}\par
\textbf{at different SNR values (shown in legend).}\par\medskip
\includegraphics[width=1.8\columnwidth]{plots/SingleSystemEta_q4_0_mc2_25_chi0_50}
\caption{We show here probability distributions for mass ratio $\eta$ as measured
for the same signal at different SNRs. The intrinsic parameters of the source
are: $q = \mbh/\mns = 5.4M_\odot/1.35M_\odot = 4$, $\chibh=+0.5$, and $\lambdans=2000$;
and the signal is injected at SNRs $\rho=\{20,30,50\}$ (left to right). The templates
{\it ignore} tidal effects.
% 
In each panel: the dashed red line marks the median or {\it measured} value
$\eta^\mathrm{Median}$, while the dashed green line show the true value
$\eta^\mathrm{Injected}$. The darker shading shows
the recovered $90\%$ credible interval for $\eta$, $(\Delta\eta)^{90\%}$.
% 
Comparing systematic and statistical errors, we find that:
at $\rho=20$, $\eta$ measurement is dominated by statistical
errors; at $\rho=30$, the two become comparable; and 
for louder signals ($\rho\simeq50$), the systematic errors dominate.
}
\label{fig:SingleSystemEtaPDFvsSNR}
\end{figure*}
% 
% ###############################
% In this section we summarize various technical aspects of this paper. The model
% used to obtain tidal waveforms is described in Sec.~\ref{s2:waveforms}. The 
% process of inferring source parameters from a GW signal is summarized in 
% Sec.~\ref{s2:bayesian}. The process of generating and combining multiple events
% is instead detailed in the corresponding results section,
% Sec.~\ref{s2:astro_multiple}.

@@

\textcolor{blue}{%
Credible intervals $(\Delta X)^{90\%}$ showing the precision with which
$X=\{\mchirp,\eta,\chibh\}$ can be measured, are presented in
Appendix~\ref{as1:nontidalerrors}.
}
We remind ourselves that this {\it precision} is only meaningful so long as the
measurement is {\it accurate} to begin with. Therefore, we define $\arr_X$ as
the ratio between systematic and statistical errors associated with the
measurement of parameter $X$,
\begin{equation}\label{eq:arr}
\arr_X = \dfrac{(X^\mathrm{Median} - X^\mathrm{Injected})}{(\Delta X)^{90\%}},
\end{equation}
in order to compare the relative magnitude of both. Only when
$|\arr_X| < 1$ can we ignore tidal effects in our templates
without hampering the measurement of non-tidal parameters from NSBH signals.
% 
We start with calculating $\arr_{\mchirp}$ as a function of various source
parameters and show it in Fig.~\ref{fig:TN_chirpMassBias_vs_Lambda_SNR}.
$\mchirp$ is the leading order
mass combination that affects the GW strain emitted by compact binaries as they
spiral in, and is therefore determined the most precisely. We notice
immediately that for $\rho\leq 30$ the systematics are well under control
and so we can obtain reliable chirp mass estimates for NSBH signals using BBH
templates.
% 
For louder and less likely SNRs ($\rho\simeq 50$), we find that
$\arr_{\mchirp}$ can become comparable to unity, but only if the BH has
prograde spin $\chibh\gtrsim 0.4$, {\it and} the true NS tidal deformability
is large enough, s.t. $\lambdans \gtrsim 1000$.
% 
We can therefore conclude that only for very loud signals, with
$\rho\gtrsim 50-70$, will the inclusion of tidal terms in our template
models improve our $\mchirp$ estimation. For lower SNRs, inclusion of new
physical content in templates will instead get washed out by detector noise.
% 
In addition, we also note that $\arr_{\mchirp}\geq 0$ always,
i.e. $\mchirp$ is always being over-estimated. This is to be expected since
the tidal deformation and disruption of NSs close to merger reduces GW signal
power at high frequencies, making the resulting signal resemble a BBH signal of
lower merger frequency - and therefore of higher mass.

@@

In the previous section, we showed that the effects of the tidal deformation of
NSs by their companion BHs become discernible in
the GW spectrum under certain favorable conditions, including (a) BH mass is
sufficiently small, (b) BH spin is positive aligned, i.e. $\chibh\gtrsim +0.4$,
(c) the NS is not very compact, with $\lambdans\gtrsim 1000$, and (d) the
source location and orientation are such that its GW SNR $\gtrsim 30$.
% 
Both condition (a) and (b) enhance the tidal distortion of the star and increase
the number of orbits the system goes through at small separation, where the
differences between NSBH and BBH signals are maximal.
% 
Conditions (a)-(c) also reduce the onset frequency of the disruption of the NS,
allowing for it to happen earlier in the orbit. 
% 
We expect that these conditions are also the ones which should maximize the
likelihood of {\it measuring} tidal effects in NSBH signals. Here,
%
%In addition, a SNR of $30$ corresponds to a $1.35+5M_\odot$ binary with 
%$\chibh = 0.4$ at a luminosity distance of $XXX$~Mpc, if optimally aligned
%with LIGO ($XXX$~Mpc if the binary subtends a $45^o$ angle with LIGO's line of
%sight.
%
%These conditions are strong, but not unrealistic. For e.g., a detection rate
%of $20$ NSBHs a year implies $1-2$ of them will have SNRs$\gtrsim 30$, since
%the number of expected events scales as $1/\rho^2$ if we assume that sources
%are distributed uniformly in spacial volume.
%
we turn the question around to ask: under similarly favorable circumstances,
can we gain insights about the internal structure of neutron stars from GW
observations?

@@

In the previous section, we showed that single observations of NSBH
coalescences at moderate SNRs have little information about the internal
structure of neutron stars that will be accessible to Advanced LIGO at its
design sensitivity. We expect all neutron stars to share the same equation of
state, and hence the same $\lambdans(\mns)$. In addition, we know that the mass
distribution of (most) NSs that have not been spun up to millisecond periods
(which are the ones we focus on in this paper, by setting $\chins\approx 0$) is
narrowly peaked around $\sim 1.35M_\odot$~\cite{Kiziltan2013}. Therefore,
information from multiple NSBH observations can be combined to improve our
estimation of $\lambdans$. We explore the same in this section within a fully
Bayesian framework. We refer the reader to Ref.~\cite{Mandel:2009pc,Lackey2014,
Wade:2014vqa} for similar analyses of BNS inspirals.

@@

multiple {\it identical} sources with realistic but different SNRs. Let us consider the case

@@

In order to generate each population, the first step is to fix
the NS properties: (i) NS mass $\mns=1.35M_\odot$, (ii) NS spin $\chins=0$
and (iii) NS tidal deformability $\lambdans=$ fixed value chosen from
$\{500,800,1000,1500,2000\}$. Next, we generate events, by sampling BH mass
(uniformly) from $\mbh\in[3M_\odot,6.75M_\odot]$, BH spin (uniformly)
from $\chibh\in[0, 1]$, orbital inclination from $\iota\in[0,\pi]$, and 
source location uniform in spatial volume\footnote{with a minimum SNR 
$\rho_\mathrm{min}=10$}.
% 
We restrict ourselves to positive aligned BH spins, since binaries with
anti-aligned spins have very little information to add at realistic SNRs,
as demonstrated in Fig.~\ref{fig:TT_LambdaCIWidths90_0_Lambda_SNR}. This is
to be taken into account when the number of observations is related to detector
operation time. 
% 
We repeat this process till we have an ordered set of $N$ events.
% 
Since we want to analyze not just a single realization of an astrophysical
population, but an ensemble of them, we make an additional approximation to
mitigate computational cost. Complete Bayesian parameter estimation is
performed for a set of simulated signals whose parameters are the vertices
of a regular hyper-cubic grid (henceforth ``G'') in the space of
$\{q\}\times\{\chibh\}\times\{\rho\}$, with each sampled at $q=\{2,3,4,5\}$,
$\chibh=\{-0.5,0,0.5,0.75\}$, and $\rho=\{10,20,30,50,70\}$.
All events in each population draw are substituted by their respective nearest
neighbours on the grid G.
% 
Our chosen signal parameter distribution is different from some other studies
in literature, which often sample from more astrophysically informed population
distribution functions~\cite{Mandel:2009pc}.

@@

First, we study the effects of not including tidal terms in our search
templates while characterizing NSBH signals. We expect that the waveform
template that best fits the signal would compensate for the reduced number of
degrees of freedom in the template model by moving away from the true
parameters of the binary. This should result in a {\it systematic} bias in 
the recovered values of non-tidal source parameters, such as its masses 
and spins. In order to quantify it, we inject tidal signals into zero noise,
and perform a Bayesian parameter estimation analysis on them using templates
{\it without} tidal terms.
% 
We use the LEA+ model (c.f. Sec.~\ref{s2:waveforms}) to produce tidal waveforms
that incorporate the effect
of NS distortion during inspiral, and of its disruption close to merger. Our
injected signals sample the region of NSBH parameter space where NS disruption
prior to binary merger is likely {\it and} can be modeled using LEA+. Their
parameters are given by combinations of $q=\mbh/\mns=\{2,3,4,5\},
\chibh=\{-0.5, 0, 0.5, 0.75\}$ and $\lambdans=\{500,800,1000,1500,2000\}$.
Other parameters, such as source location and orientation, that factor out of
$h(t)$ as amplitude scaling are co-sampled by varying $\rho=\{20,30,50,70\}$.

@@

% \prayush{\bf Sec 5 summary}\\
% \prayush{\bf Tidal measurements versus observation time - in how many years can
% we constrain $\lambdans$ reasonably?}\\
% % 
Third, knowing that single observations can furnish only so much information
about the NS equation of state, we move on to investigate how well we do with
multiple signals. In order to quantify how $\lambdans$ measurement improves
with the number of observed events $N$, we generate populations of NSBH signals
and combine the information extracted from each event.
% 
The population generation procedure is as follows. The neutron star mass is
held fixed at $1.35M_\odot$, its spin at $\chins=0$, and its tidal
deformability is fixed to each of $\lambdans=\{500,1000,1500,2000\}$. Black hole
mass is sampled uniformly from the range $[2,5]\times 1.35=[2.7, 6.75]M_\odot$,
and spin from $\chibh\in[0,1]$. As before, our parameter choice here is given
by the intersection set of the mass range that allows for neutron star disruption
and the range supported by LEA+~\cite{Foucart2012,Foucart:2013a,Lackey:2013axa}.
In order to keep the computational cost reasonable, we make an additional
approximation. For every population generated, we replace the parameters of each
event by their nearest neighbor on the uniform grid G, which has vertices
at: $q=\{2,3,4,5\}\times\chibh=\{-0.5,0,0.5,0.75\}\times\lambdans=\{500,800,
1000,1500,2000\}\times\rho=\{10,20,30,50,70\}$.
% 
This way, we only have to run full Bayesian parameter estimation analysis on
this fixed set of signals. 
% 
There are two sources of error that enter the deductions we make from
a single population generated in the manner described above. First, since the
injection parameters are pushed to their nearest neighbor on a grid, we
find discrete jumps in $\lambdans$ errors as a function of $N$. And second, an
individual population is one particular realization of a stochastic process and
could have excursions that may never be found in another population. To
account for both of these limitations, we generate an ensemble of populations,
and conservatively combine information from all of them\footnote{See 
Sec.~\ref{s1:multiple_observations} for further details.}.

@@

Finally, we note that the underlying numerical simulations used to calibrate
the waveform model used here have not been verified against
independent codes so far.
% 
It is therefore difficult to assess the combined modeling error of LEA+ and its
effect
on our results. Our results here are, therefore, limited by the limitations of
our waveform model, and presented with this caveat. However, we do expect the
combined effect of modeling errors to {\it not} affect our {\it qualitative}
conclusions.
% 
In future, we plan to further the results presented here by using more recent
tidal models~\cite{Pannarale:2015jka,Hinderer:2016eia}, that
may improve upon LEA+\footnote{One of them~\cite{Pannarale:2015jka} is only an
amplitude model though, which has to be augmented with a compatible phase model
first.}.
